What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-09 android-stacks.tex
new content to undocumented section

Manualy inspecting the code of the methods throwing direct instances of RuntimeException revealed 
another and even worse bad practice:  none of the analyzed methods 
documented this exception as part of the exception interface (e.g., using @throws annotation).

Such undocumented runtime exceptions represent a serious threat to system robustness, specially
when such exceptions are thrown then the third party code is invoked inside the application (e.g. libraries, or framework utility code).
In such cases the client usualy do not have access to the source code, and in the absense of
the exception documentation it is very difficult or even impossible for the client of such thrid party code to 
protect the application against undocumented runtime exceptions. As a consequenve, the
 undocumented runtime exception may remain uncaught and lead to system crashes.

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. If we filter the exceptions implicitly signaled by 
JVM (due to programming mistakes) the number of runtime exceptions thrown by third party code is XXX 
signaled by Android and YYY signaled by library code.

@@

Manually inspecting 10% of each set we could also observe that none of the methods throwing 
runtime exceptions documented the exceptions on the method signature (as suggested by Bloch
 ~\cite{bloch2008effective}) nor included a comment in Javadoc. This result is aligned with the results of other 
study conducted by from Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

On the other hand, we could observe that the checked exceptionsmost represent most of the  
exceptions signaled by the set of libraries reused by Android platform (i.e., org.apache.harmony,
org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax.) referred in
this work as libcore. It is pointed as a good practice
for libraries (see Section ~\ref{best}) because by using checked exceptions
libraries can define a precise exception interface to its clients. Such
libraries are heavily used in several projects, and such precise exception
interface may be related to the libraries maturity.

>>>>

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. After filtering all the exceptions implicitly signaled by 
JVM (due to programming mistakes) and inspecting the signaler methods of such exceptions (i.e.,  direct an inderect 
instances of java.lang.RuntimeException). We could observe observe that only 1 method out of 118 inspected signalers
 (i.e., 0,8%) documented the explicitly thrown runtime exceptions in a Javadoc comment, and none of the analized 
included the exception  as part of the exception interface of the method (i.e., using throws clause on method signature). 
This result is aligned with the results of other  study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

Such undocumented runtime exceptions represent a serious threat to system robustness, specially
when such exceptions are thrown then the third party code is invoked inside the application (e.g. libraries, or framework utility code).
In such cases the client usualy do not have access to the source code, and in the absense of
the exception documentation it is very difficult or even impossible for the client of such thrid party code to 
protect the application against undocumented runtime exceptions. As a consequence, the
 undocumented runtime exception may remain uncaught and lead to system crashes.

@@

On the other hand, we could observe that most of the exceptions signaled by the set of 
libraries reused by Android platform (i.e., org.apache.harmony,
org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax - referred in
this work as libcore) were checked exceptions. It is pointed as a good practice
for libraries (see Section ~\ref{best}) because by using checked exceptions
libraries can define a precise exception interface to its clients. Such
libraries are heavily used in several projects, and such precise exception
interface may be related to the libraries maturity.
